---
title: "Homework 1"
author: "Kelsey Grinde and Brian Williamson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

## Instructions

This assignment should be submitted on Canvas as a .pdf or Word document by 12:30pm on March 30th. You are welcome to work together on the homework; however, final solutions should be written up individually. Unedited software output in the body of your assignment is not acceptable. Please round numerical values to 2--3 significant digits, and include your R code (if applicable) in a .R file uploaded as a separate document.

If you give an honest attempt for all of the questions, you will receive 2 points for completeness. The remaining 8 points are based on correctness, as specififed in the problem statements.

## 1. Create an .Rmd file (2.5 points)

We will use `R` and `RStudio` extensively throughout the quarter, so for this homework assignment we want to make sure that you are able to generate a PDF or Word document containing your homework solutions using `R Markdown` and the `knitr` package. 

To complete your assignment, please complete the following steps: 

1. Download `hw1.Rmd` from Canvas.

2. Open this file in RStudio: start RStudio > File > Open File > find `hw1.Rmd` in your files

3. Update the author: on the third line of the file, change the author name from "Kelsey Grinde and Brian Williamson" to "Your Name"

4. Enter your answers to Questions 2 and 3. 

5. Install the knitr package if you haven't done so already.

6. Once you have completed all steps above, create a PDF or Word document: click the **Knit** button at the top of this screen.

To get credit for this problem, please turn in this `.Rmd` file along with  the final PDF or Word document you generated.

## 2. Study design (1.5 points)

For each of the following scientific questions, propose an appropriate study design. (You may use the same design at most twice.) Provide a one-sentence explanation of your reasoning. 

### a. Among people who are already heavy smokers, can running reduce the effect of smoking on incidence of lung cancer? 

### b.	How does risk of severe influenza differ between men and women?

### c. Is a common genetic mutation associated with colon cancer? 


## 3. Background survey (4 points)

We want to know a little more about you! Please answer the following questions. Credit for this question will be assigned entirely on the basis of completion.

### a. What is your preferred name?


### b. What is your year and major?


### c. What are your career and/or post-graduation goals?


### d. What previous statistics/biostatistics courses have you taken and when? 


### e. Do you have any other experience with statistics (e.g., job or research)?


### f. Do you have any previous computing experience (e.g., Excel, R, Stata, Matlab)? If so, in what context?


### g. Do you have a dataset in mind that you would like to use for the course project? If so, please briefly describe here. 


### h. Why are you taking this course? What are you hoping to learn from this class? 


### i. What is one thing you enjoy doing during your free time?


### j. Is there anything else you would like us to know about you?


## R Code

```{r define-file-names, include = FALSE}
##### STEP 1: Update these file names to reflect the names of this R Markdown file (rmd.name) and the name of the R file where you want to send your R code (code.name). #####
rmd.name <- 'hw1.Rmd' # name of this file
code.name <- 'hw1.R' # name of R code file
```

```{r extract-code, include = FALSE, eval = FALSE}
##### STEP 2: Once you have completed your assignment, run these commands #####
library('knitr') 
purl(input = rmd.name, output = code.name) # send your R code to code.name 
```

Please upload the file `r code.name` along with your completed assignment. 
