---
title: "Homework 2"
author: "Kelsey Grinde and Brian Williamson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

# Instructions

This assignment should be submitted on Canvas as a .pdf or Word document by 12:30pm on the due date. You are welcome to work together on the homework; however, final solutions should be written up individually. 

Make sure you update the document header to reflect your name and the type of document you want to create (pdf_document or word_document). Unedited software output in the body of your assignment is not acceptable. Please round numerical values to 2--3 significant digits, and include your R code in a .R file uploaded as a separate document. See the "How to Write a Proper Homework" document on Canvas for additional guidelines.  

If you give an honest attempt for all questions, you will receive 2 points for completeness. The remaining 8 points are based on correctness, as specififed in the problem statements.

# 1. FEV: descriptive statistics and t-tests (5.5 points) 
Forced expiratory volume (FEV) is a measure of lung capacity. Higher values indicate higher lung capacity (the ability to blow out more air in one second), and therefore indicate better lung function. In this problem, you will use R to perform an analysis to examine the difference in mean FEV between smokers and nonsmokers, among children ages 3-19. 

## a. (0.5 points) Explain in one sentence why you cannot use a randomized controlled trial to answer this question about the relationship between smoking and FEV in kids.

Type your answer to this problem here.

## b. (0 points) Load the FEV dataset into R.
Download the FEV dataset from Canvas (Files > Data > fev.txt). Move the fev.txt file **to the folder where this .Rmd document is saved** (otherwise the code below won't work). Then, load the dataset by running the following code chunk (either highlight lines XX--XX and click the "Run" button at the top right of this *pane*, or click the "Play" button at the top right corner of the *chunk*).

If you've done everything correctly, you should see this:

```{r code-for-prob-1b, include = TRUE}
## load the FEV data
fev <- read.delim(file = "fev.txt", sep = "", header = TRUE)
## check that you've read it in correctly:
head(fev)
```

**(No answer required for this problem.)**

## c. (0.5 points) Create a table of descriptive statistics in R.
We want to characterize the distribution of FEV overall and stratified by smoking status. We will use some functions from the "uwIntroStats" package to do this.

First, make sure you've installed the "uwIntroStats" package: run the command `install.packages("uwIntroStats")`). 

Now, run the following code to create a table of descriptive statistics:

```{r code-for-prob-1c, eval = FALSE}
# load uwIntroStats package
library('uwIntroStats')
# create descrip table for FEV
descrip(fev$fev, strata=fev$smoke)
```

*(You do not need to report the table, just run the commands and look at the result so you can answer the following questions. Note that in this dataset, "smoke = 1" denotes smokers and "smoke = 2" denotes nonsmokers.)*

**Looking at the table you created, how does lung capacity differ between the two groups? Is this result surprising?**

Type your answer to this problem here.

## d. (0.5 points) Provide a histogram of FEV for smokers and nonsmokers, separately. Use the histograms to qualitatively characterize the difference in the distribution of FEV between smokers and nonsmokers.

To do this, run the following commands:

```{r code-for-prob-1d, fig.height=3, fig.width=6, eval = FALSE}
# tell R to arrange plots in one row, 2 columns
par(mfrow=c(1,2)) 
# histogram of fev among smokers (smoke==1)
hist(subset(fev,smoke==1)$fev, xlab='FEV (l/sec)', main="Histogram (smokers)") 
# histogram of fev among nonsmokers (smoke==2)
hist(subset(fev,smoke==2)$fev, xlab='FEV (l/sec)', main="Histogram (nonsmokers)") 
```

**Please include your histograms in your report:** to do this, you simply need to change the statement `eval = FALSE` to `eval = TRUE` at the end of line XX and the histograms will automatically appear in your .pdf or Word document once you hit the Knit button! *(Try this now and make sure it works.)*

**What differences do you see between smokers and nonsmokers in the histograms you created? Is there evidence of unequal variances between groups?**

Type your answer to this problem here.

## e. (0.5 points) Perform an analysis to assess whether the mean FEV in smokers is different from the mean FEV in nonsmokers.

Perform a $t$-test which presumes equal variances between groups. Run the following code:

```{r code-for-prob-1e, eval=FALSE}
ttest(fev$fev, by = fev$smoke, var.eq = TRUE)
```

**Briefly state and interpret the results (difference in means between the two groups, 95% confidence interval, p-value, conclusion).**

Type your answer to this problem here.

## f. (0.5 points) Repeat problem (e), except with the test that allows for unequal variances between smokers and nonsmokers. 

This can be done with the following command:

```{r codde-for-prob-1f, eval=FALSE}
ttest(fev$fev, by = fev$smoke, var.eq = FALSE)
```

**Briefly state and interpret the results (difference in means between the two groups, 95% confidence interval, p-value, conclusion).**

Type your answer to this problem here.

## g. (1 point) Does your estimate of the mean difference change when you allow for unequal variances between groups? Does the confidence interval change? Do your scientific conclusions change?

Type your answer to this problem here.

## h. (1 point) Which analysis---that of (e) or (f)---would you prefer? Explain your answer in a sentence or two. What if you were not permitted to use your responses to problems (c) and (d) to justify your answer?

Type your answer to this problem here.

## i. (0.5 points) In either of your analyses, do you reach the conclusion that smoking increases lung capacity? Explain.

Type your answer to this problem here.

## j. (0.5 points) If you answered "No" to the previous part, restate the conclusion so that it is accurate. Then, provide a logical explanation for why you may have observed this somewhat unexpected phenomenon. Note: for this problem, you will be evaluated on effort only. 

Type your answer to this problem here.

## 2. (YY points) (HW2, Q1)

Type your answer to this problem here.

```{r code-for-prob-2, include = FALSE}
## Fill this in with code executing the R commands you need for this problem
## Comment your code so that you (and us, if necessary) know what you're doing
```


## 3. (ZZ points) (HW2, Q2)

# R Code

Once you have completed your assignment, update and run the `define-file-names` and `extract-code` code chunks to extract the R code from this .Rmd file and send it to a .R file. 

```{r define-file-names, include = FALSE}
##### STEP 1: Update these file names #####
rmd.name <- 'hw2.Rmd' # name of this file; update if you changed the name of this file
code.name <- 'hw2.R' # name of R code file; you can change if desired
```

```{r extract-code, eval = FALSE, include = FALSE}
##### STEP 2: Run these commands #####
library('knitr') 
purl(input = rmd.name, output = code.name) # send your R code to code.name 
```

Please upload the file `r code.name` along with your completed assignment. 