---
title: "Homework 8"
author: "Kelsey Grinde and Brian Williamson"
date: "21 May 2018"
output: pdf_document
urlcolor: blue
---

# Instructions

This assignment should be submitted on Canvas as a .pdf or Word document **by 5:00pm on June 1**. You are welcome to work together on the homework; however, final solutions should be written up individually. 

Make sure you update the document header to reflect your name and the type of document you want to create (pdf_document or word_document). **Unedited R output, or code, in the body of your assignment is not acceptable.** Please round numerical values to 2--3 significant digits, and include your R code in a .R file uploaded as a separate document. See the "How to Write a Proper Homework" document on Canvas for additional guidelines.  

You will receive full credit for this assignment if you attempt Problem 1 and your choice of Problem 2 or Problem 3 below. However, please do attempt all problems prior to the exam, as these questions are similar to the types of analyses that you will do in the future!

# 1. Review the most recent version of the course lecture notes. Please find at least 4 typos, and write down: (a) which slide the typo is on, (b) why you think it is a typo, and (c) the corrected version of the relevant part of the slide (just give us enough to go off of so that we can find the typo easily).

\newpage

# 2. Cerebral atrophy and mortality in US senior citizens

A government sponsored cohort study of adults aged 65 years and older was conducted to observe the incidence of cardiovascular disease (especially heart attacks and congestive heart failure) and cerebrovascular disease (especially strokes) in the elderly over an 11 year period, and to relate the incidence of those diseases to various risk factors measured in the population on a regular basis. These data contain information on 735 generally healthy adults taken from the Medicare rolls. After being followed for some time, the participants were asked to submit to magnetic resonance imaging (MRI) of their brains. 

In imaging older people, it has been observed that brains tend to atrophy (shrink) with age, but it has not yet been established whether these changes should be viewed as part of the normal aging process, whether these changes are merely signs of other disease such as heart disease, or whether these changes are indicative of separate disease processes. 

In this problem, you will investigate the association between **cerebral atrophy** and **mortality**. Please consider only the following additional variables:

* *age:* the participant's age at the time of MRI, in years 
* *male:* an indicator of the participant's sex (1 = male, 0 = female)
* *race:* the participant's race (1 = white, 2 = black, 3 = Asian, 4 = other)
* *packyrs:* participant smoking history in pack years (1 pack year = smoking 1 pack of cigarettes per day for 1 year; a participant who has never smoked has 0 pack years)
* *alcoh:* average alcohol intake for the participant for the two weeks prior to MRI (drinks per week, where 1 drink = 1 oz. whiskey, 4 oz. wine, or 12 oz. beer)
* *stroke:* indicator of whether the participant had been diagnosed with a cerebrovascular event prior to MRI (0 = no, 1 = transient ischemic attack, 2 = stroke. Cerebrovascular disease refers to the narrowing of blood vessels that supply the brain. In mild cases, the patient may experience a short period of weakness or paralysis, and then recover completely [transient ischemic attack]. In severe cases, a portion of the brain is deprived of oxygen and dies [a stroke])
* *diabetes:* indicator of whether the participant had been diagnosed with diabetes prior to MRI (1 = yes, 0 = no)
* *dsst:* score on a test of cognitive function (0 = low, 100 = high)
* *atrophy:* measure of global brain atrophy; 0 = no atrophy, 100 = most severe atrophy
* *obstime:* the total time (in days) that the participant was observed on-study from date of MRI to death or September 16, 1997, whichever came first
* *death:* indicator of death (1 = yes, 0 = no)

Download the MRI dataset from Canvas (`Files > Data > mri.txt`) prior to running any statistical analyses. 

## a. Your scientific goal is to investigate the relationship between cerebral atrophy (predictor of interest) and death (outcome). Which covariates would you adjust for as potential confounders and precision variables? Justify your answer (1 sentence per variable). You may include a causal diagram to help with your justification.

## b. Create a new variable, `death_in_5`, that indicates whether or not the participant died within 5 years of their MRI. Is this a valid way to define the outcome, death? If so, why? If not, why not? Provide descriptive statistics to justify your answer.

## c. Create boxplots of cerebral atrophy score (on the y-axis) versus death within 5 years of MRI (x-axis). What does this plot suggest?

## d. Fit a linear regression of death within 5 years of MRI on cerebral atrophy, adjusted for the potential confounders and precision variables that you identified in (a). Interpret the coefficient estimate for your parameter of interest.

## e. Fit a logistic regression of death within 5 years of MRI on cerebral atrophy, adjusted for the potential confounders and precision variables that you identified in (a). Interpret the coefficient estimate for your parameter of interest.

## f. Fit a Poisson regression of death within 5 years of MRI on cerebral atrophy, adjusted for the potential confounders and precision variables that you identified in (a). Interpret the coefficient estimate for your parameter of interest.

## g. Which of the above analyses [linear regression (d), logistic regression (e), or Poisson regression (f)] would you have chosen prior to looking at the data, if your collaborator approached you with this scientific question?

## h. Now create a survival object that combines time to death and the event indicator of death. Plot estimated survival curves for time to death, stratified by high (>= 50) versus low (< 50) atrophy score. Interpret the survival curves: what is the median survival time in each group? Do the curves differ?

## i. Fit a Cox proportional hazards regression of time to death on cerebral atrophy, adjusted for the potential confounders and precision variables that you identified in (a). Interpret the coefficient estimate for your parameter of interest.

## j. Which of the above analyses [linear regression (d), logistic regression (e), Poisson regression (f), or Cox proportional hazards regression (i)] would you have chosen prior to looking at the data, if your collaborator approached you with this scientific question?

## k. Compare and contrast the results you obtained across the four regression analyses. Did you observe anything surprising?

## l. Discuss which assumptions are necessary for statistical inference based on your chosen analysis in (j) to be valid.

\newpage

# 3. Complications of pregnancy in South Africa

Over the past two centuries there have been enormous improvements in the incidence of perinatal morbidity and mortality (i.e., infant illness and death in the weeks just prior to and just after birth). Many of these gains have been attributed to the increase in adequate prenatal care that women receive during pregnancy. A major focus of that care is in the prevention of early delivery and low birthweight of babies: In one classification of United States data in a 2005 report, medical problems attributable to short gestation and low birthweight are the second leading cause of infant mortality (after congenital malformations).

As might be expected, perinatal mortality remains a large problem in much of the developing world where adequate prenatal care is not as available: In the 2005 report from the World Health Organization it was noted that 1 in 5 African women lose a baby during their lifetime, as opposed to the 1 in 125 rate in richer countries. In less developed countries, many urban pregnant women receive their only care from nurses in busy antenatal community clinics, where they will typically also deliver their babies. A challenge in this environment is to develop low technology methods of identifying problem pregnancies so that these can be appropriately referred to a higher level of care in a timely manner. Some problem outcomes of concern are low birth weight babies (<2500 gms) as might be caused by either pre-term delivery (prior to the 38th week of gestation) or small for gestational age (SGA) babies (below the 10th percentile of birth weight for the gestational age at which they are born).

In this problem, you will investigate the association between **maternal age** and **birthweight**. Additional variables include:

* *ht:* the mother's height in cm
* *age:* the mother's age at enrollment, in years
* *sga:* small for gestational age (1 = yes, 0 = no)
* *parity:* number of prior deliveries
* *smoker:* indicator of mother's smoking status (1 = yes, 0 = no)
* *bweight:* birthweight of infant in grams
* *sex:* sex of the infant (1 = male, 2 = female)
* *gesage:* gestational age at delivery (weeks)

Download the pregnancy dataset from Canvas (`Files > Data > pregout.txt`) prior to running any statistical analyses. 

## a. Your scientific question is to investigate the relationship between maternal age and birthweight. Which covariates would you adjust for as potential confounders and precision variables? Justify your answer (1 sentence per variable). You may include a causal diagram to help with your justification.

## b. Create a scatterplot of maternal age (on the x-axis) against birthweight (on the y-axis). What does this plot tell you about the relationship between birthweight and maternal age?

## c. Fit a linear regression of birthweight on maternal age, adjusted for the potential confounders and precision variables you identified in (a). Interpret the coefficient estimate for your parameter of interest.

## d. Create a new variable, `lbw` (1 = low birthweight [< 2500 g], 0 = not low birthweight). What proportion of babies are low birthweight in this sample?

## e. Fit a linear regression of low birthweight on maternal age, adjusted for the potential confounders and precision variables you identified in (a). Interpret the coefficient estimate for your parameter of interest.

## f. Fit a logistic regression of low birthweight on maternal age, adjusted for the potential confounders and precision variables you identified in (a). Interpret the coefficient estimate for your parameter of interest.

## g. Fit a Poisson regression of low birthweight on maternal age, adjusted for the potential confounders and precision variables you identified in (a). Interpret the coefficient estimate for your parameter of interest.

## h. Which of the above analyses [linear regression with continuous birth weight as outcome (c), linear regression with binary low birth weight as outcome (e), logistic regression with binary low birthweight as outcome (f), or Poisson regression with binary low birth weight as outcome (g)] would you have chosen prior to looking at the data, if your collaborator approached you with this scientific question?

## i. Compare and contrast the results you obtained across the four regression analyses. Did you observe anything surprising?

## k. Discuss which assumptions are necessary for statistical inference based on your chosen analysis in (h) to be valid.

\newpage

# R Code

Once you have completed your assignment, update (if needed) and run the `define-file-names` and `extract-code` code chunks to extract the R code from this .Rmd file and send it to a .R file. 

```{r define-file-names, include = FALSE}
##### STEP 1: Update these file names #####
rmd.name <- 'hw8.Rmd' # name of this file; update if you changed the name of this file
code.name <- 'hw8.R' # name of R code file; you can change if desired
```

```{r extract-code, eval = FALSE, include = FALSE}
##### STEP 2: Run these commands #####
library('knitr') 
purl(input = rmd.name, output = code.name) # send your R code to code.name 
```

**Please upload the file `r code.name` along with your completed assignment.**
