---
title: "Homework"
author: "Kelsey Grinde and Brian Williamson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

## Instructions

This assignment should be submitted on Canvas as a .pdf or Word document by 12:30pm on the due date. You are welcome to work together on the homework; however, final solutions should be written up individually. 

Make sure you update the document header to reflect the assignment number (e.g., "Homework 1"), your name ("Kelsey Grinde and Brian Williamson" > "Your Name"), and the type of document you want to create (pdf_document or word_document). Unedited software output in the body of your assignment is not acceptable. Please round numerical values to 2--3 significant digits, and include your R code (if applicable) in a .R file uploaded as a separate document. See the "How to Write a Proper Homework" document on Canvas for additional guidelines.  

If you give an honest attempt for all questions, you will receive 2 points for completeness. The remaining 8 points are based on correctness, as specififed in the problem statements.

## 1. (XX points)

Type your answer to this problem here.

```{r code-for-prob-1, include = FALSE}
## This is a sample code chunk! 
## Fill this in with code executing the R commands you need for this problem
## Comment your code so that you (and us, if necessary) know what you're doing
```

## 2. (YY points)

Type your answer to this problem here.

```{r code-for-prob-2, include = FALSE}
## Fill this in with code executing the R commands you need for this problem
## Comment your code so that you (and us, if necessary) know what you're doing
```

## R Code

(If you did not need any R code to complete this assignment, skip this.)

Once you have completed your assignment, update and run the `define-file-names` and `extract-code` code chunks to extract the R code from this .Rmd file and send it to a .R file. 

```{r define-file-names, include = FALSE}
##### STEP 1: Update these file names #####
rmd.name <- 'hw_template.Rmd' # name of this file
code.name <- 'hw_template.R' # name of R code file
```

```{r extract-code, eval = FALSE, include = FALSE}
##### STEP 2: Run these commands #####
library('knitr') 
purl(input = rmd.name, output = code.name) # send your R code to code.name 
```

Please upload the file `r code.name` along with your completed assignment. 
