---
title: "Discussion Section 8"
author: "Kelsey Grinde and Brian Williamson"
date: "15 May 2018"
output: 
  pdf_document:
    fig_caption: true
---

# Instructions

This activity should be submitted on Canvas as a .pdf or Word document **by 12:00pm on May 15**. You should work in groups of 3--4 to complete this assignment, but each person should turn in his/her own activity to Canvas at the end of class.

Make sure you update the document header to reflect your name and the type of document you want to create (pdf_document or word_document). **Unedited R output, or code, in the body of your assignment is not acceptable.** Please round numerical values to 2--3 significant digits.

# Data

For this activity, we're going to look at the `leukemia` data you're analzying on Homework 7. Here's a brief summary of the study:

- **Type of study:** randomized controlled trial (RCT)
- **Scientific goal:** evaluate whether an experimental chemotherapy regimen improves survival rates in leukemia patients (compared to the standard of care)
    - **Outcome:** survival time
    - **Predictor of interest:** treatment group
- **Time period:** enrollment starts on March 26, 2015 and study ends 3 years later, on March 26, 2018
- **Sample size:** 1,000 participants (500 randomized to the experimental treatment and 500 to the standard of care)

Download the data from `Canvas > Data > leukemia.txt`. The dataset contains the following variables:

- `tx`: treatment assignment (1 = experimental therapy, 0 = standard of care)
- `obs_time`: observation time (time in days until death or censoring)
- `event`: event indicator (1 = death, 0 = censoring)
- `death_in_6_months`: indicator of whether or not the subject died within 6 months of study enrollment (1 = died in 6 months, 0 = still alive after 6 months)

Read the dataset into `R` using the `read-data` code chunk below. You may need to update the file name to reflect where you saved the data.

```{r read-data,echo=F,eval=T}
# load data into R
# you may need to update this to reflect the correct file location
leuk <- read.table('leukemia.txt', header = TRUE)
```

# 1. Create a survival object

To work with time-to-event data in `R`, we need to create a variable of a specific type, called a survival object.  

The first step is to load the `survival` package. This package comes automatically installed so we do not need to install it first; we only need to use `library` to load it.  

```{r load-survival,echo=F,warning=F,message=F}
# load the survival package
library('survival')
```

Once the `survival` package is loaded, we can use one of its functions (the `Surv` function) to create a time-to-death variable. Add this new time-to-death variable to your data frame.

```{r create-ttodeath,echo=F,eval=T}
# add ttodeath to dataset
leuk$ttodeath <- Surv(time = leuk$obs_time, event = leuk$event)
```

Look at the first few rows of your dataset (hint: use the `head` function), and compare the original `obs_time` variable to the new `ttodeath` variable. 

```{r look-at-data,echo=F,eval=F}
# enter code to look at dataset here

```

**Why do some of the `ttodeath` values contain a plus sign?**

Type your answer here.

# 2. Plot and interpret Kaplan-Meier curves

Now, you are going to plot the Kaplan-Meier estimate of the survival curves among those in the treatment group and those in the standard of care group. 

First we will create a `survfit` object, which creates the survival curves.

```{r create-curves, echo = F, eval = F}
# create a survfit object (estimates the KM curve)
surv_obj <- survfit(ttodeath ~ tx, data = leuk)
```

Then we plot the curves using the familiar `plot` function. (Make sure to change the chunk options to `eval = T` so that your plot will show up in your knitted document.)

```{r kaplan-meier, echo = F, eval = F}
# plot the survival curves
plot(surv_obj, lwd = 2, # lwd makes lines thicker
     col = c('blue', 'red'), # blue for tx=0, red for tx=1
     xlab = 'Days of follow-up', 
     ylab = 'Proportion surviving')

# add a legend to the top right of the plot
legend('topright', legend = c('Standard Care', 'Treatment'), lty = 1, col = c('blue', 'red'))
```

## i. Which group has a higher mean survival time? State your reasoning.

Type your answer here.

## ii. Can you estimate the median survival time? If so, provide the relevant estimate(s). If not, why not?

Type your answer here.

## iii. Do you think this plot suggests that the treatment is effective? Explain your reasoning.

Type your answer here.

# 3. Run and interpret a logrank test

Now, we will perform a logrank test to test the null hypothesis that the two survival curves (for the standard of care and treatment groups) are equal.

Run the following code chunk to perform the test.

```{r logrank,echo=F,eval=F}
## logrank test
survdiff(ttodeath ~ tx, data = leuk)
```

**Report the results of this test (i.e., provide the p-value and your statistical and scientific conclusions).**

Type your answer here.

\newpage

# Bonus Question: proportional hazards regression

Now, perform a proportional hazards regression with time-to-death as the outcome and treatment group as the predictor. To fit this regression model, run the following code chunk.

```{r ph-regression,echo=F,eval=F}
# load uwIntroStats package
library(uwIntroStats)

# fit proportional hazards regression model
regress("hazard", ttodeath ~ tx, data = leuk)
```

**Report your results in 4--5 sentences.**

Type your answer here.
