---
title: "Introduction to R and RStudio"
subtitle: "BIOST 311, Discussion Section 01"
author: "Kelsey Grinde and Brian Williamson"
date: "27 March 2018"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## function to color text
colFmt = function(x,color){
  outputFormat = knitr::opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\textcolor{",color,"}{",x,"}",sep="")
  else if(outputFormat == 'html')
    paste("<font color='",color,"'>",x,"</font>",sep="")
  else
    x
}
```

## R: the big picture

R is a `r colFmt("free", "blue")`, `r colFmt("open source", "green")` software package that can be used for data analysis, graphics, and programming.

At its core, R is an interactive, command-driven `r colFmt("language", "blue")`: you type a command, R executes the command, and then returns results.

While it may take some time to develop a mastery of R, it is relatively easy to get set up with the basics and start analyzing data.

Learning R is worth your time, and has many advantages, including:

  * R is free and open source
  * there is an active group of contributors to R software
  * you can do flexible things with your data
  * many people have written packages that make data analysis easier

## RStudio: the interface to R

While R is the `r colFmt("language", "blue")` we will use to analyze data, it does not come packaged with the best `r colFmt("interface", "green")`. An interface connects you to the program: for example, your operating system lets you point and click on icons (and read things easily) rather than writing in 0's and 1's, the language that the computer understands.

Instead of using the R interface packaged with the download, we will use RStudio. This is also a `r colFmt("free", "blue")`, `r colFmt("open source", "green")` software package, but is only an interface to R---this means that even if we type into RStudio, the commands will be executed in R.

RStudio adds many things over the basic R interface, including:

  * An improved layout of tools for data analysis
  * Support for embedding reproducible research tools into data analysis
  * Support for building your own R packages
  * Integrated R help and documentation
  * De-bugging tools
  
The combination of R and RStudio makes it easy to:

  * write R scripts containing all code for an analysis, along with comments describing what the code does
  * write reports with code embedded (in Rmarkdown, more on this later)
  * organize your data analysis workflow (e.g., reading in data, accessing help files)

When you open RStudio for the first time, the window will display three `r colFmt("panels", "blue")`---these have different functions, as we'll now discover.

### The console

The default position of the console is in the upper right-hand pane of RStudio. This pane is our real window into R: any commands that we want R to execute must be entered in the console, and any results that are output by these commands, with the exception of plots, will appear in the console.

To enter a command into the console, move your cursor to the `>` and click. When there is a flashing `|` in the console, you're ready to enter a command. From now on, we'll call the `>` the execution line. Once you've entered a command on the execution line, hit `Enter` to run the command.

For example, we could type in the number `47` and hit enter, which prints out the number `47`:
```{r example-47, include = TRUE, echo = TRUE}
47
```

### The script editor

### The environment/history pane

### The files/plots/packages/help pane

## R: commands

### Functions

### Creating objects

### Reading in a dataset

### Manipulating datasets

## Getting help

## R packages
