---
title: "Code for Chapter 1, section 2 (Multiple Linear regression)"
author: "Kelsey Grinde and Brian Williamson"
date: "April 12, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=400)
```

```{r read-fev, echo = TRUE, include = FALSE}
# fev <- read.table('http://www.emersonstatistics.com/datasets/fev.txt',header=T)
fev <- read.table('../../datasets/fev.txt', header = TRUE)
fev$smoker <- ifelse(fev$smoke == 2, 0, 1)
adultfev <- read.table("adultfev.txt", header = TRUE)
adultfev <- adultfev[complete.cases(adultfev), ]
# infarcts <- read.table("http://www.emersonstatistics.com/datasets/infarcts.txt", header = TRUE)
```

## FEV on smoking, stratified by height
```{r fev-smoke-height-strat, echo = TRUE}
library("uwIntroStats", warn.conflicts = FALSE)
less_60 <- subset(fev, height < 60)
greater_60 <- subset(fev, height >= 60)
regress("mean", fev ~ smoker, data = less_60)
```
\newpage
```{r }
library("uwIntroStats", warn.conflicts = FALSE)
less_60 <- subset(fev, height < 60)
greater_60 <- subset(fev, height >= 60)
regress("mean", fev ~ smoker, data = greater_60)
```

\newpage

## FEV on smoking, adjusting for binary height
```{r echo = TRUE}
library("uwIntroStats", warn.conflicts = FALSE)
regress("mean", fev ~ smoker + I(height >= 60), data = fev)
```

\newpage

## FEV on smoking, adjusting for continuous height
```{r echo = TRUE}
library("uwIntroStats", warn.conflicts = FALSE)
regress("mean", fev ~ smoker + height, data = fev)
```

\newpage

## FEV on smoking, adjusting for age, height, sex (adult FEV)
```{r echo = TRUE}
library("uwIntroStats", warn.conflicts = FALSE)
regress("mean", fev ~ smoke + age + male + height, data = adultfev)
```

\newpage

## FEV on smoking, adjusted for age, sex (adult FEV)
```{r echo = TRUE}
library("uwIntroStats", warn.conflicts = FALSE)
regress("mean", fev ~ smoke + age + male , data = adultfev)
```

\newpage

## FEV with EM between smoking and sex
```{r echo = TRUE}
library("uwIntroStats", warn.conflicts = FALSE)
regress("mean", fev ~ smoke*male, data = adultfev)
```

\newpage

## DSST, EM with age and educ
```{r echo = TRUE, eval = FALSE}
library("uwIntroStats", warn.conflicts = FALSE)
regress("mean", dsst ~ age*educ, data = infarcts)
```

## FEV prediction
```{r echo = TRUE}
library("uwIntroStats", warn.conflicts = FALSE)
regress("mean", fev ~ smoke*male + height, data = adultfev)
```

\newpage

## FEV prediction, part 2
```{r echo = TRUE}
mod_predict_fev <- lm(fev ~ smoke*male + height, data = adultfev)
head(cbind(adultfev, fitted(mod_predict_fev)))
predict(mod_predict_fev, newdata = data.frame(smoke = 0, male = 0, height = 60))
predict(mod_predict_fev, newdata = data.frame(smoke = 1, male = 1, height = 72))
```