---
title: "Discussion Section 6"
author: "Kelsey Grinde and Brian Williamson"
date: "1 May 2018"
output: 
  pdf_document:
    fig_caption: true
---

# Instructions

This assignment should be submitted on Canvas as a .pdf or Word document by 12:00pm on May 1. You should work in groups of 2--3 to complete this assignment.

Make sure you update the document header to reflect your name and the type of document you want to create (pdf_document or word_document). **Unedited R output, or code, in the body of your assignment is not acceptable.** Please round numerical values to 2--3 significant digits, and include your R code in a .R file uploaded as a separate document. See the "How to Write a Proper Homework" document on Canvas for additional guidelines.  

If you give an honest attempt for all questions, you will receive 2 points for completeness. The remaining 8 points are based on correctness, as specififed in the problem statements.

# Re-creating the analyses from the Chapter 1 overview slides

In this study, over 5,000 elderly, generally healthy (cancer was
an exclusion criterion), adults were randomly selected from
Medicare rolls in four communities.

For this analysis, we are interested in the role of inflammation
in the pathogenesis of atherosclerotic disease. In particular,
we are interested in two biochemical markers of
inflammation, the C reactive protein (CRP) and
fibrinogen.

We have the following information on each subject: age
(years), sex (`male`: male = 1, female = 0), cholesterol
(`cholest`, mg/dL), BMI (`bmi`, kg/m 2 ), previous history of
cardiovascular disease (`prevdis`: 1/0), fibrinogen (`fib`,
mg/dL), and CRP (`crp`, mg/L).

Answer the following two scientific questions:

1. Are blood levels of fibrinogen associated with blood levels of CRP?
2. If blood levels of fibrinogen are associated with blood levels of CRP, does this association depend on cholesterol level?

Include at least on table or figure to support each of your answers to these statistical questions. Additionally, include a full summary of the results for each question. You are encouraged to reproduce the results found in slides 1.172--1.196.

\newpage

# R Code

Once you have completed your assignment, update (if needed) and run the `define-file-names` and `extract-code` code chunks to extract the R code from this .Rmd file and send it to a .R file. 

```{r define-file-names, include = FALSE}
##### STEP 1: Update these file names #####
rmd.name <- 'discussion6.Rmd' # name of this file; update if you changed the name of this file
code.name <- 'discussion6.R' # name of R code file; you can change if desired
```

```{r extract-code, eval = FALSE, include = FALSE}
##### STEP 2: Run these commands #####
library('knitr') 
purl(input = rmd.name, output = code.name) # send your R code to code.name 
```

**Please upload the file `r code.name` along with your completed assignment.**
